\documentclass{article}
\usepackage{mathrsfs}
\usepackage{amsmath, amssymb, amsthm, geometry, mathtools}
\usepackage{hyperref}
\usepackage{enumitem}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{remark}{Remark}[section]

\title{Twisted K-Theory and Fredholm Operators}
\author{Maksym Sokhatskyi}
\date{May 10, 2025}

\begin{document}

\maketitle

\begin{abstract}
This article explores the mathematical foundations of topological quantum programming, focusing on the role of twisted complex K-theory (\(\mathrm{KU}^\tau_G(X)\)) and the space of self-adjoint, odd-graded Fredholm operators (\(\text{Fred}^0_C\)) as presented in the framework of Twisted Equivariant Differential K-theory (TED-K). We provide a detailed explanation of the type-theoretic definition of \(\mathrm{KU}^\tau_G(X)\) in cohesive homotopy type theory, its categorical interpretation, and its significance in encoding anyonic quantum states. Additionally, we analyze the properties of \(\text{Fred}^0_C\), its connection to \(L^2\) spaces, and its role in K-theory via index theory and integration. The article is motivated by the TED-K framework for hardware-aware quantum programming, as developed by Sati and Schreiber.
\end{abstract}

\tableofcontents

\section{Twisted K-Theory and Fredholm Operators}

Topological quantum computations (TQC) leverages topological
properties of quantum materials to achieve fault-tolerant
quantum computing, primarily through the braiding of anyons
in topologically ordered states \cite{Nayak2008}.
The TED-K framework, introduced by Sati and Schreiber \cite{SatiSchreiber2022},
proposes a novel approach to topological quantum programming that is hardware-aware,
reflecting the physics of anyonic ground states via twisted equivariant differential
K-theory (TED-K). Central to this framework are the twisted
K-theory group \(\mathrm{KU}^\tau_G(X)\) and the space of
Fredholm operators \(\text{Fred}^0_C\), which encode quantum
states and their braiding operations.

This article synthesizes detailed explanations of \(\mathrm{KU}^\tau_G(X)\) and \(\text{Fred}^0_C\), drawing from the TED-K framework. We elucidate their definitions in cohesive homotopy type theory (HoTT), their categorical interpretations in \(\infty\)-topoi, and their physical significance in TQC. The exposition is aimed at readers with a background in algebraic topology, functional analysis, or quantum computing, seeking to understand the mathematical underpinnings of topological quantum programming.

\subsection{Definition in Cohesive HoTT}

In the TED-K framework, the twisted complex K-theory group \(\mathrm{KU}^\tau_G(X)\) is defined in cohesive homotopy type theory as a dependent type:

\begin{equation}
X: \text{Type}, \tau: X \to \mathrm{BPU} \vdash \mathrm{KU}^\tau_G(X) \equiv \left| \int_{\mathrm{BPU}} \left( X \to \text{Fred}_C^0 \beta \mathrm{PU} \right) \right|_0 : \text{Type}
\end{equation}

Here, \(X\) is a type representing a topological space or orbifold (e.g., the configuration space of anyonic defects), \(\tau\) is a twist map to the classifying space \(\mathrm{BPU}\), \(\text{Fred}_C^0\) is the space of self-adjoint, odd-graded Fredholm operators, \(\beta \mathrm{PU}\) denotes the homotopy fiber adjusting for the twist, \(\int_{\mathrm{BPU}}\) is the shape modality over \(\mathrm{BPU}\), and \(|\cdot|_0\) is the 0-truncation modality.

\subsection{Components of the Definition}

\begin{itemize}
    \item \textbf{\(\mathrm{PU}\)}: The projective unitary group, defined as \(\mathrm{PU} = \mathrm{U}(\mathscr{H}) / \mathrm{U}(1)\), where \(\mathrm{U}(\mathscr{H})\) is the unitary group of a separable complex Hilbert space \(\mathscr{H}\). It acts on \(\text{Fred}_C^0\) by conjugation and has \(\pi_3(\mathrm{PU}) \cong \mathbb{Z}\), making it suitable for classifying twists.
    \item \textbf{\(\mathrm{BPU}\)}: The classifying space of \(\mathrm{PU}\), with \(\pi_4(\mathrm{BPU}) \cong \mathbb{Z}\). A map \(\tau: X \to \mathrm{BPU}\) classifies a gerbe or projective bundle, encoding topological obstructions in twisted K-theory.
    \item \textbf{\(\text{Fred}_C^0\)}: The space of self-adjoint, odd-graded Fredholm operators with index 0, discussed in detail in Section \ref{sec:fredholm}.
    \item \textbf{\(\text{Fred}_C^0 \beta \mathrm{PU}\)}: The homotopy fiber of the \(\mathrm{PU}\)-action map \(\text{Fred}_C^0 \to \mathrm{BPU}\), ensuring compatibility with the twist \(\tau\).
    \item \textbf{\(X \to \text{Fred}_C^0 \beta \mathrm{PU}\)}: The type of sections of the twisted bundle of Fredholm operators over \(X\).
    \item \textbf{\(\int_{\mathrm{BPU}}\)}: The shape modality, extracting the homotopy type, combined with the dependent product over \(\mathrm{BPU}\), computing all sections compatible with \(\tau\).
    \item \textbf{\(|\cdot|_0\)}: The 0-truncation, yielding a set (the K-theory group \(\mathrm{KU}^0(X, \tau)\)).
\end{itemize}

\subsection{\(\mathrm{PU}\): Projective Unitary Group}
\label{subsec:pu}

The projective unitary group \(\mathrm{PU}\) is defined as the quotient:

\begin{equation}
\mathrm{PU} = \mathrm{U}(\mathscr{H}) / \mathrm{U}(1)
\end{equation}

where \(\mathrm{U}(\mathscr{H})\) is the group of unitary operators on a separable complex Hilbert space \(\mathscr{H}\) (e.g., \(L^2(\mathbb{R}^n; \mathbb{C})\)), equipped with the strong operator topology, and \(\mathrm{U}(1) \cong S^1\) is the circle group acting by scalar multiplication. An element \(u \in \mathrm{U}(\mathscr{H})\) is unitary if \(u^* u = u u^* = I\), preserving the inner product, and \(\mathrm{PU}\) identifies unitaries differing by a phase \(e^{i\theta} \in \mathrm{U}(1)\).

The topological structure of \(\mathrm{PU}\) is significant for K-theory, as it has non-trivial homotopy groups, notably:

\begin{equation}
\pi_3(\mathrm{PU}) \cong \mathbb{Z}
\end{equation}

This makes \(\mathrm{PU}\) suitable for classifying twists in twisted K-theory, as twists are often associated with elements in the third cohomology group \(H^3(X; \mathbb{Z})\). In the TED-K framework, \(\mathrm{PU}\) acts on \(\text{Fred}_C^0\) by conjugation:

\begin{equation}
T \mapsto u T u^{-1}, \quad u \in \mathrm{PU}, T \in \text{Fred}_C^0
\end{equation}

This action preserves the self-adjointness, odd-graded property, and Fredholm nature of the operators, enabling the encoding of the twist \(\tau\). Physically, the \(\mathrm{PU}\)-action reflects symmetries in the quantum system, such as those arising from gauge fields or anyonic statistics in topological quantum materials.

In the categorical setting of the cohesive \(\infty\)-topos of smooth \(\infty\)-groupoids, \(\mathrm{PU}\) is a group object (a 0-type with a group structure). Its role is to mediate the twisting of K-theory classes, ensuring that the quantum states encoded by \(\text{Fred}_C^0\) respect the topological phase specified by \(\tau\).

\subsection{\(\mathrm{BPU}\): Classifying Space of \(\mathrm{PU}\)}
\label{subsec:bpu}

The classifying space \(\mathrm{BPU}\) is the delooping of \(\mathrm{PU}\), a 1-type in the \(\infty\)-topos characterized by:

\begin{equation}
\pi_1(\mathrm{BPU}) \cong \mathrm{PU}, \quad \pi_i(\mathrm{BPU}) = 0 \text{ for } i \neq 1
\end{equation}

However, since \(\mathrm{PU}\) itself has higher homotopy groups, \(\mathrm{BPU}\) inherits non-trivial homotopy in higher degrees:

\begin{equation}
\pi_4(\mathrm{BPU}) \cong \pi_3(\mathrm{PU}) \cong \mathbb{Z}
\end{equation}

Maps \(\tau: X \to \mathrm{BPU}\) classify principal \(\mathrm{PU}\)-bundles or gerbes over \(X\), corresponding to elements in \(H^1(X; \mathrm{PU})\), which are related to \(H^3(X; \mathbb{Z})\) via the long exact sequence of the fibration:

\begin{equation}
\mathrm{U}(1) \to \mathrm{U}(\mathscr{H}) \to \mathrm{PU}
\end{equation}

In the TED-K framework, \(\tau: X \to \mathrm{BPU}\) encodes a topological obstruction, such as a gauge field or a gerbe, that affects the braiding statistics of anyons. The classifying space \(\mathrm{BPU}\) serves as the base for the twist, and the homotopy fiber \(\text{Fred}_C^0 \beta \mathrm{PU}\) ensures that the Fredholm operators are compatible with the \(\mathrm{PU}\)-bundle defined by \(\tau\).

Categorically, \(\mathrm{BPU}\) is a pointed 1-type in the cohesive \(\infty\)-topos, and the map \(\tau\) is a morphism in \(\mathcal{H}\). The dependent product \(\iint_{\mathrm{BPU}}\) in the definition of \(\mathrm{KU}^\tau(X)\) integrates over all possible twists, ensuring that the K-theory classes reflect the topological structure imposed by \(\tau\). Physically, \(\mathrm{BPU}\) mediates the connection between the topological phase of the quantum material and the quantum states encoded in \(\text{Fred}_C^0\).


\subsection{Categorical Interpretation}

The definition is interpreted in the \(\infty\)-topos of smooth \(\infty\)-groupoids, a cohesive \(\infty\)-topos \(\mathcal{H}\) equipped with modalities like the shape functor \(\int: \mathcal{H} \to \text{HoTop}\). The type \(\mathrm{KU}^\tau_G(X)\) corresponds to the homotopy classes of sections of a twisted bundle in the slice \(\infty\)-topos \(\mathcal{H}_{/\mathrm{BPU}}\). Categorically:

\begin{itemize}
    \item The mapping space \(X \to \text{Fred}_C^0 \beta \mathrm{PU}\) is the object \(\text{Map}(X, \text{Fred}_C^0 \beta \mathrm{PU})\) in \(\mathcal{H}\).
    \item The dependent product \(\int_{\mathrm{BPU}}\) is the right adjoint to the base change along \(\tau\), and \(\int\) extracts the homotopy type.
    \item The 0-truncation \(|\cdot|_0\) maps to the set of connected components, yielding the abelian group \(\mathrm{KU}^\tau_G(X)\).
\end{itemize}

\subsection{Physical Significance}

In TQC, \(\mathrm{KU}^\tau_G(X)\) encodes the **ground states of su(2)-anyons** (e.g., Majorana or Fibonacci anyons) in topological quantum materials \cite{SatiSchreiber2022Anyonic}. The configuration space \(X\) represents defect positions, and the twist \(\tau\) accounts for gauge fields or gerbes affecting anyonic statistics. The **braid group**, arising from the cohesive shape \(\int X\), acts on \(\mathrm{KU}^\tau_G(X)\) via transport in HoTT, implementing quantum gates through adiabatic braiding. This makes \(\mathrm{KU}^\tau_G(X)\) a hardware-aware construct, enabling formal verification and classical simulation of quantum computations \cite{SatiSchreiber2022}.

\subsection{Fredholm Operators: \(\text{Fred}^0_C\)}
\label{sec:fredholm}

\subsubsection{Definition of Fredholm Operators}

A **Fredholm operator** \(T: \mathscr{H} \to \mathscr{H}\) on a complex Hilbert space \(\mathscr{H}\) is a bounded linear operator with:

\begin{enumerate}
    \item Finite-dimensional kernel: \(\dim(\ker T) < \infty\).
    \item Finite-dimensional cokernel: \(\dim(\text{coker } T) = \dim(\mathscr{H} / \text{im } T) < \infty\).
    \item Closed image: \(\text{im } T\) is closed in \(\mathscr{H}\).
\end{enumerate}

The index is:

\begin{equation}
\text{index}(T) = \dim(\ker T) - \dim(\text{coker } T)
\end{equation}

In \(\text{Fred}^0_C\), operators are **self-adjoint**, **odd-graded**, and typically have **index 0**.

\subsubsection{Properties of \(\text{Fred}^0_C\)}

\begin{itemize}
    \item \textbf{Self-Adjointness}: \(T = T^*\), where \(T^*\) is the adjoint satisfying \(\langle T x, y \rangle = \langle x, T^* y \rangle\). This ensures a real spectrum and is common in spectral K-theory.
    \item \textbf{Odd-Graded}: \(\mathscr{H} = \mathscr{H}^0 \oplus \mathscr{H}^1\) is \(\mathbb{Z}/2\)-graded, with a grading operator \(\gamma: \mathscr{H} \to \mathscr{H}\), \(\gamma^2 = I\), and \(T\) satisfies \(\gamma T = -T \gamma\). Thus, \(T = \begin{pmatrix} 0 & T_1 \\ T_2 & 0 \end{pmatrix}\).
    \item \textbf{Index 0}: \(\dim(\ker T) = \dim(\text{coker } T)\), corresponding to the degree-0 component of K-theory.
    \item \textbf{\(\mathrm{PU}\)-Action}: The projective unitary group \(\mathrm{PU}\) acts by conjugation, \(T \mapsto u T u^{-1}\), preserving the Fredholm properties and enabling twisted K-theory.
\end{itemize}

\subsubsection{Connection to \(L^2\) Spaces}

The Hilbert space \(\mathscr{H}\) is often an \(L^2\) space, such as \(L^2(\mathbb{R}^n; \mathbb{C})\), the space of square-integrable functions:

\[
L^2(\mathbb{R}^n; \mathbb{C}) = \left\{ f: \mathbb{R}^n \to \mathbb{C} \mid \int_{\mathbb{R}^n} |f(x)|^2 \, dx < \infty \right\}
\]

Fredholm operators on \(L^2\) spaces include elliptic differential operators, e.g., \(T = -\Delta + V\), where \(\Delta\) is the Laplacian and \(V\) is a potential. For grading, consider \(\mathscr{H} = L^2(\mathbb{R}^n; \mathbb{C}) \oplus L^2(\mathbb{R}^n; \mathbb{C})\), with \(\gamma = \begin{pmatrix} I & 0 \\ 0 & -I \end{pmatrix}\). Operators in \(\text{Fred}^0_C\) act on this graded space, encoding anyonic wavefunctions in TQC.

\subsubsection{Integration and Index Theory}

The analytical properties of \(\text{Fred}^0_C\) involve integration, particularly in **index theory**. The index of a Fredholm operator can be computed via the Atiyah-Singer index theorem:

\begin{equation}
\text{index}(T) = \int_M \text{ch}(E) \wedge \text{Td}(TM)
\end{equation}

where \(\text{ch}(E)\) is the Chern character and \(\text{Td}(TM)\) is the Todd class. In differential K-theory, the Chern character of a K-theory class involves integrals of curvature forms:

\begin{equation}
\text{ch}([T]) = \int_X \text{tr}(e^{F/2\pi i})
\end{equation}

In cohesive HoTT, the shape modality \(\int\) acts as a homotopical integration, mapping \(\mathrm{KU}^\tau_G(X)\) to topological invariants, crucial for computing braiding statistics in TQC.

\subsubsection{Physical Role}

\(\text{Fred}^0_C\) operators act on the Hilbert space of anyonic wavefunctions, with self-adjointness and grading reflecting their symmetries and statistics. The \(\mathrm{PU}\)-action enables twisting, and the index 0 condition aligns with the degree-0 K-theory group, classifying stable quantum states. Integration via the Chern character connects these operators to physical observables, facilitating simulation and verification in the TED-K framework.

\subsection{Conclusion}

The twisted K-theory group \(\mathrm{KU}^\tau_G(X)\) and the space of Fredholm operators \(\text{Fred}^0_C\) are foundational to the TED-K framework for topological quantum programming. \(\mathrm{KU}^\tau_G(X)\), defined in cohesive HoTT, encodes anyonic ground states and supports braid quantum gates via the braid group’s action. \(\text{Fred}^0_C\), consisting of self-adjoint, odd-graded Fredholm operators on \(L^2\) spaces, provides the analytical backbone, with integration playing a key role in index theory and differential K-theory. Together, these constructs enable a hardware-aware, formally verifiable approach to TQC, bridging topology, functional analysis, and quantum computing. For further details, see \cite{SatiSchreiber2022Supplementary}.

\begin{thebibliography}{9}
\bibitem{Nayak2008} C. Nayak et al., Non-Abelian anyons and topological quantum computation, \emph{Rev. Mod. Phys.} 80 (2008), 1083--1159.
\bibitem{SatiSchreiber2022} H. Sati and U. Schreiber, Topological Quantum Programming in TED-K, \emph{ncatlab.org/schreiber/show/TQCinTEDK}, 2022.
\bibitem{SatiSchreiber2022Anyonic} H. Sati and U. Schreiber, Anyonic topological order in TED-K-theory, \emph{arXiv:2206.13563}, 2022.
\bibitem{SatiSchreiber2022Supplementary} H. Sati and U. Schreiber, Supplementary material for TQC in TED-K, \emph{ncatlab.org/schreiber/show/TCinTEDKAGMConAbs}, 2022.
\end{thebibliography}

\end{document}
